Chemically Self-Consistant Modeling of the Globular Cluster NGC 2808 and its Effects on the Inferred Helium abundance of Multiple Stellar Populations.

### 2023-12-10 src/ms.tex
Incorperated Brian's and Greg's notes

@@

\begin{abstract}
  The Helium abundances in the multiple populations which are now known to
  comprise all closley studies Milky Way globular clusters are often inferred
  by fitting isochohrones generated from from stellar evolutionary models to
  globular cluster photometry. It is therefore important to understand build
  stellar models that are chemically self-consistent in the compositions of the
  structure, atmosphere, and opacity. In this work we present the first
  chemically self-consistent stellar models of the Milky Way Globular Cluster
  NGC 2808 using MARCS model atmospheres, OPLIB high-temperature radiative
  opacities, and AESOPUS low-temperature radiative opacities. In order to to
  this we develope a robust software suite for globular cluster photometry,
  \fidanka, which can be used for both population synthethis as well as
  isochrone optimization. \fidanka's primary goal is to determine, in a
  relativly unbiased way, the ideal number of populations within photometric
  dataset to fit isochrones to. We achive this through a combinarion of Baysian
  Guassian Mixture Modeling and a novel number density estimation algorithm.
  Using \fidanka and F275W-F814W photometry from the Hubble UV Globular Cluster
  Survey we find that the helium abundance of the second generation of stars in
  NGC 2808 is higher than the first generation by $15\pm3\%$. This is in
  agreement with previous studies of NGC 2808.
\end{abstract}
% \begin{abstract}
% 	Globular Clusters (GCs) provide a unique astrophysical laboratory for
% 	studying the formation and evolution of stars. GCs are old, dense, and it
% 	has historically been believed that they have a single stellar population. However, in the
% 	last two decades, it has been definitively shown that most if not all Milky
% 	Way GCs have multiple stellar populations (MPs). These MPs are chemically
% 	distinct from one another, primarily separated by light element, including helium, abundance
% 	variations without the standard accompanying heavy element abundance
% 	variations. As the precise formation channel of these MPs remains
% 	an open question, and one which is sensitive to the population - population
% 	compositional differences, the extent of the composition
% 	variations between MPs is a key parameter to constrain. Many metal abundances may be
% 	directly measured spectroscopically; however, helium abundances are not
% 	directly observable in GCs. Instead, helium abundances are inferred from
% 	stellar models. It is therefore important to understand build stellar
% 	models that are self-consistent in the compositions of the structure,
% 	atmosphere, and opacity. In this work we present the first chemically
% 	self-consistent stellar models of the Milky Way Globular Cluster NGC 2808 using MARCS model atmospheres, OPLIB high-temperature radiative opacities, and AESOPUS low-temperature radiative opacities.
% 	We find that the helium abundance of the second generation of stars is
% 	higher than the first generation by {\color{red} SOME AMOUNT}.
% 	{\color{blue} This is in agreement with previous studies of NGC 2808.}
% \end{abstract}

>>>>

\author{Amanda Ash}
\affiliation{Department of Physics and Astronomy, University of North Georgia, Dahlonega, GA 30533, USA}

@@

\begin{abstract}
  The Helium abundances in the multiple populations which are now known to
  comprise all closely studied Milky Way globular clusters are often inferred
  by fitting isochohrones generated from stellar evolutionary models to
  globular cluster photometry. It is therefore important to build stellar
  models that are chemically self-consistent in terms of their structure,
  atmosphere, and opacity. In this work we present the first chemically
  self-consistent stellar models of the Milky Way Globular Cluster NGC 2808
  using MARCS model atmospheres, OPLIB high-temperature radiative opacities,
  and AESOPUS low-temperature radiative opacities. These stellar models were
  fit to the NGC 2808 photometry using \fidanka, a new software tool that was
  developed optimally fit cluster photometry to isochrones and for population
  synthesis.  \fidanka can determine, in a relatively unbiased way, the ideal
  number of distinct populations which exist within a dataset and then fits
  isochrones to each population. We achieve this through a combination of
  Bayesian Gaussian Mixture Modeling and a novel number density estimation
  algorithm. Using \fidanka and F275W-F814W photometry from the Hubble UV
  Globular Cluster Survey we find that the helium abundance of the second
  generation of stars in NGC 2808 is higher than the first generation by
  $15\pm3\%$. This is in agreement with previous studies of NGC 2808.
\end{abstract}
% \begin{abstract}
% 	Globular Clusters (GCs) provide a unique astrophysical laboratory for
% 	studying the formation and evolution of stars. GCs are old, dense, and it
% 	has historically been believed that they have a single stellar population. However, in the
% 	last two decades, it has been definitively shown that most if not all Milky
% 	Way GCs have multiple stellar populations (MPs). These MPs are chemically
% 	distinct from one another, primarily separated by light element, including helium, abundance
% 	variations without the standard accompanying heavy element abundance
% 	variations. As the precise formation channel of these MPs remains
% 	an open question, and one which is sensitive to the population - population
% 	compositional differences, the extent of the composition
% 	variations between MPs is a key parameter to constrain. Many metal abundances may be
% 	directly measured spectroscopically; however, helium abundances are not
% 	directly observable in GCs. Instead, helium abundances are inferred from
% 	stellar models. It is therefore important to understand build stellar
% 	models that are self-consistent in the compositions of the structure,
% 	atmosphere, and opacity. In this work we present the first chemically
% 	self-consistent stellar models of the Milky Way Globular Cluster NGC 2808 using MARCS model atmospheres, OPLIB high-temperature radiative opacities, and AESOPUS low-temperature radiative opacities.
% 	We find that the helium abundance of the second generation of stars is
% 	higher than the first generation by {\color{red} SOME AMOUNT}.
% 	{\color{blue} This is in agreement with previous studies of NGC 2808.}
% \end{abstract}

### 2023-12-10 src/sections/fidanka.tex
Incorperated Brian's and Greg's notes

\section{fidanka}\label{sec:fidanka}
When fitting isochrones to the data we have four main criteria for any method

\begin{itemize}
	\item The method must be robust enough to work along the entire main sequence, turn off, and much of the subgiant and red giant branchs.
	\item Any method should consider photometric uncertainty in the fitting process.
	\item The method should be model independent, weighting any n number of populations equally.
	\item The method should be automated and require minimal intervention from the user.
\end{itemize}

We do not believe that any currently available software is a match for
our use case. Therefore, we elect to develop our own software suite, \fidanka.
\fidanka is a python package designed to automate much of the process of
measuring fiducial lines in CMDs, adhering to the four criteria we lay out
above. Primary features of \fidanka may be separated into three
categories: fiducial line measurement, stellar population synthesise, and
isochrone optimization/fitting. Additionally, there are utility functions which
are detailed in the \fidanka documentation.

\subsection{Fiducial Line Measurement}
\fidanka takes a iterative approach to measuring fiducial lines, the first step
of which is to make a ``guess'' as to the fiducial line. . This initial guess
is calculated by splitting the CMD into magnitude bins, with uniform numbers of
stars per bin (so that bins are cover a small magnitude range over densely
populated regions of the CMD while covering a much larger magnitude range in
sparsely populated regions of the CMD, such as the RGB). A unimodal Gaussian
distribution is then fit to the color distribution of each bin, and the
resulting mean color is used as the initial fiducial line guess. This rough
fiducial line will approximately trace the area of highest density. The initial
guess will be used to verticalze the CMD so that further algorithms can work in
1-D magnitude bins without worrying about weighting issues caused by varying
projections of the evolutionary sequence onto the magnitude axis.
Verticalization is preformed taking the difference between the guess fiducial
line and the color of each star in the CMD.

If \fidanka were to simply apply the same algorithm to the verticalized CMD
then the resulting fiducial line would likely be a re-extraction of the initial
fiducial line guess. To avoid this, we take a more robust, number density based
approach, which considers the distribution of stars in both color and magnitude
space simultaneously. For each star in the CMD we first using a
\texttt{introselect} partitioning algorithm to select the 50 nearest stars in F814W vs. F275W-F814W space.
To account for the case where the star is at an extreme edge of the CMD, those
50 stars include the star itself (such that we really select 49 stars + 1). We
use \texttt{qhull}\footnote{https://www.qhull.com}\citep{Barber1996, } to
calculate the convex hull of those 50 points. The number density at each star
then is defined as $50/A_{hull}$, where $A_{hull}$ is the area of the convex
hull. Because we use a fixed number of points per star, and a partitioning
algorithm as opposed to a sorting algorithm, this method scales like
$\mathcal{O}(n)$, where n is the number of stars in the CMD. This method also
intrinsically weights the density of of each star equally as the counting
statistics per bin are uniform. We are left with a CMD where each star
has a defined number density (Figure \ref{fig:densityMapDemo}).

@@

\fidanka can now exploit this density map to fit a better fiducial line to the
data, as the density map is far more robust to outliers. There are multiple
algorithms we implement to fit the fiducial line to the color-density profile
in each magnitude bin (Figure \ref{fig:densityBinsDemo}); they are explained in more detail in the \fidanka
documentation. However, of most relevance here is the Bayesian Gaussian Mixture
Modeling (BGMM) method. BGMM is a clustering algorithm which, for some fixed
number of n-dimensional Gaussian distributions, $K$, determines the mean, covariance, and
mixing probability (somewhat analogous to amplitude) of each $k^{th}$
distribution, such that the local lower bound of the evidence of each star
belonging strongly to a single distribution is maximized.

@@

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{Notebooks/Figures/vertFit.png}
	\caption{CMD where points are colored by density. Line trace the infered 
	fiducial line(s) in each magnitude bin.}
	\label{fig:vertFit}
\end{figure}

@@

We can adapt this density map based BGMM method to consider photometric
uncertainties by adopting a simple Monte Carlo approach. Instead of measuring
the fiducial line(s) a single time, \fidanka can measure the fiducial line(s)
many times, resampling the data with replacement each time. For each resampling
\fidanka adds a random offset to each filter based on the photometric
uncertainties of each star. From these $n$ measurments the mean fiducial line
for each sequence can be identified along with upper and lower bound confidence
intervals in each magnitude bin.

\subsection{Stellar Population Synthesis}
In addition to measuring fiducial lines, \fidanka also includes a stellar
population synthesise module. This module is used to generate synthetic CMDs
from a given set of isochrones. This is of primary importance for binary
population modelling. The module is also used to generate synthetic CMDs for
the purpose of testing the fiducial line extraction algorithms against priors.

>>>>

\section{fidanka}\label{sec:fidanka}
When fitting isochrones to the clusters with multiple populations we have four
main criteria for any method

\begin{itemize}
  \item The method must be robust enough to work along the entire main
    sequence, turn off, and much of the subgiant and red giant branch.
	\item Any method should consider photometric uncertainty in the fitting process.
	\item The method should be model independent, weighting any n number of populations equally.
	\item The method should be automated and require minimal intervention from the user.
\end{itemize}

We do not believe that any currently available software is a match for
our use case. Therefore, we elect to develop our own software suite, \fidanka.
\fidanka is a python package designed to automate much of the process of
measuring fiducial lines in CMDs, adhering to the four criteria we lay out
above. Primary features of \fidanka may be separated into three
categories: fiducial line measurement, stellar population synthesise, and
isochrone optimization/fitting. Additionally, there are utility functions that
are detailed in the \fidanka documentation.

\subsection{Fiducial Line Measurement}
\fidanka takes a iterative approach to measuring fiducial lines, the first step
of which is to make a ``guess'' as to the fiducial line. This initial guess
is calculated by splitting the CMD into magnitude bins, with uniform numbers of
stars per bin (so that bins are cover a small magnitude range over densely
populated regions of the CMD while covering a much larger magnitude range in
sparsely populated regions of the CMD, such as the RGB). A unimodal Gaussian
distribution is then fit to the color distribution of each bin, and the
resulting mean color is used as the initial fiducial line guess. This rough
fiducial line will approximately trace the area of highest density. The initial
guess will be used to verticalze the CMD so that further algorithms can work in
1-D magnitude bins without worrying about weighting issues caused by varying
projections of the evolutionary sequence onto the magnitude axis.
Verticalization is preformed taking the difference between the guess fiducial
line and the color of each star in the CMD.

If \fidanka were to simply apply the same algorithm to the verticalized CMD
then the resulting fiducial line would likely be a re-extraction of the initial
fiducial line guess. To avoid this, we take a more robust, number density based
approach, which considers the distribution of stars in both color and magnitude
space simultaneously. For each star in the CMD we first using an
\texttt{introselect} partitioning algorithm to select the 50 nearest stars in
F814W vs. F275W-F814W space. To account for the case where the star is at an
extreme edge of the CMD, those 50 stars include the star itself (such that we
really select 49 stars + 1). We use
\texttt{qhull}\footnote{https://www.qhull.com}\citep{Barber1996} to calculate
the convex hull of those 50 points. The number density at each star then is
defined as $50/A_{hull}$, where $A_{hull}$ is the area of the convex hull.
Because we use a fixed number of points per star, and a partitioning algorithm
as opposed to a sorting algorithm, this method scales like $\mathcal{O}(n)$,
where n is the number of stars in the CMD. This method also intrinsically
weights the density of of each star equally as the counting statistics per bin
are uniform. We are left with a CMD where each star has a defined number
density (Figure \ref{fig:densityMapDemo}).

@@

\fidanka can now exploit this density map to fit a better fiducial line to the
data, as the density map is far more robust to outliers. There are multiple
algorithms we implement to fit the fiducial line to the color-density profile
in each magnitude bin (Figure \ref{fig:densityBinsDemo}); they are explained in
more detail in the \fidanka documentation. However, of most relevance here is
the Bayesian Gaussian Mixture Modeling (BGMM) method. BGMM is a clustering
algorithm which, for some fixed number of n-dimensional Gaussian distributions,
$K$, determines the mean, covariance, and mixing probability (somewhat
analogous to amplitude) of each $k^{th}$ distribution, such that the local
lower bound of the likelyhood of each star belonging strongly to a single
distribution is maximized.

@@

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{Notebooks/Figures/vertFit.png}
	\caption{CMD where points are colored by density. Lines trace the infered 
	fiducial line(s) in each magnitude bin.}
	\label{fig:vertFit}
\end{figure}

@@

We can adapt this density map based BGMM method to consider photometric
uncertainties by adopting a simple Monte Carlo approach. Instead of measuring
the fiducial line(s) a single time, \fidanka can measure the fiducial line(s)
many times, resampling the data with replacement each time. For each resampling
\fidanka adds a random offset to each filter based on the photometric
uncertainties of each star. From these $n$ measurements the mean fiducial line
for each sequence can be identified along with upper and lower bound confidence
intervals in each magnitude bin.

\subsection{Stellar Population Synthesis}
While not extensively used in this paper \fidanka can, in addition to measuring fiducial
lines, preform stellar population synthesise. \fidanka's population synthesis
module can generate synthetic stellar population from a set of MIST formatted
isochrones. This is of primary importance for binary population modeling. The
module is also used to generate synthetic CMDs for the purpose of testing the
fiducial line extraction algorithms against priors.

### 2023-12-10 src/sections/fitting.tex
Incorperated Brian's and Greg's notes

\section{Isochrone Fitting}\label{sec:isoFit}
We fit pairs of isochrones to the HUGS data for NGC 2808 using \fidanka, as
descrbed in \S {\color{red}FIDANKA SECTION}. Two isochrones, one for Population
A and one for Population E are fit simultaneously. These isochrones are
constrained to have distance modulus, $\mu$, and color excess, E(B-V) which
agree to within 1\%. Moreover, we constrain the mixing length, $\alpha_{ML}$, for any two isochrones in a set to be within 0.5 of one and other. For every isochrone in the
set of combination of which fulfilling these constraints $\mu$, $E(B-V)$,
Age$_{A}$, and $Age_{B}$ are optimized to reduce the $\chi^{2}$ distance
between the fiducial lines and the isochrones. Because we fit fiducial lines
directly, we do not need to consider the binary population fraction, $f_{bin}$,
as a free parameter.

The best fit isochrones are shown in Figure \ref{fig:BestFitResults} and optimized
parameters for these are presented in Table \ref{tab:BestFitResults}. We find helium mas fractions which are consistent with those identified in past literature \citep[e.g.][]{Milone2015}. Note that our helium mass fraction gird has a spacing of 0.03 between grid points and we are therefore unable to resolve between certain proposed helium mass fractions for the younger sequence (for example between 0.37 and 0.39).

\begin{figure*}
  \centering
  \includegraphics[width=0.9\textwidth]{src/figures/BestFitResults.pdf}
  \label{fig:BestFitResults}
  \caption{Best fit isochrone results for NGC 2808.}
\end{figure*}

\begin{table*}
  \centering
  \begin{tabular}{c | c c c c c c}
    \hline
    population & age & distance modulus & extinction & Y & $\alpha_{ML}$ & $\chi^{2}_{\nu}$\\
    & [Gyr] & & [mag] & & &\\
    \hline
    \hline
    A & 12.3 & 14.91 & 0.54 & 0.24 & 1.901 & 0.014\\
    E & 14.3 & 14.96 & 0.54 & 0.39 & 1.750 & 0.017 \\
    \hline
  \end{tabular}
  \label{tab:BestFitResults}
  \caption{Best fit parameters derived from fitting isochrones to the fiducual lines derived from the NCG 2808 photometry.}
\end{table*}

@@

\subsection{The Number of Populartions in NGC 2808}
In order to estimate the number of populations which ideally fit the NGC 2808
F275W-F814W photometry without overfitting the data we make use of silhouette
analysis \citep[][and in a similar manner to how \citet{Valle2022}
preform their analysis of spectroscopic data]{ROUSSEEUW198753}. We find the average silhouette score for all tagged
clusters identified using BGMM in all magnitude bins over the CMD using the
standar python module \texttt{sklearn}. Figure \ref{fig:clusterAn} shows the
silhouette analysis results and that 2 populations fit the photometry most
ideally. This is in line with what our BGMM model predicts for the majority of
the the CMD.

>>>>

\section{Isochrone Fitting}\label{sec:isoFit}
We fit pairs of isochrones to the HUGS data for NGC 2808 using \fidanka, as
descrbed in \S \ref{sec:fidanka}. Two isochrones, one for Population A and one
for Population E are fit simultaneously. These isochrones are constrained to
have distance modulus, $\mu$, and color excess, E(B-V) which agree to within
0.5\% and an ages which agree to within 1\%. Moreover, we constrain the mixing length, $\alpha_{ML}$, for any two
isochrones in a set to be within 0.5 of one and other. For every isochrone in
the set of combination of which fulfilling these constraints $\mu$, $E(B-V)$,
Age$_{A}$, and $Age_{B}$ are optimized to reduce the $\chi^{2}$ distance
($\chi^{2} = \sum\sqrt{\Delta \text{color}^{2} + \Delta \text{mag} ^{2}}$)
between the fiducial lines and the isochrones. Because we fit fiducial lines
directly, we do not need to consider the binary population fraction, $f_{bin}$,
as a free parameter.

The best fit isochrones are shown in Figure \ref{fig:BestFitResults} and
optimized parameters for these are presented in Table \ref{tab:BestFitResults}.
We find helium mass fractions that are consistent with those identified in past
literature \citep[e.g.][]{Milone2015}. Note that our helium mass fraction grid
has a spacing of 0.03 between grid points and we are therefore unable to
resolve between certain proposed helium mass fractions for the younger sequence
(for example between 0.37 and 0.39).

\begin{figure*}
  \centering
  \includegraphics[width=0.9\textwidth]{src/figures/BestFitResults.pdf}
  \label{fig:BestFitResults}
  \caption{Best fit isochrone results for NGC 2808. The best fit population A
  and E models are shown as black lines. The following 50 best fit models are
  presented as grey lines. The solid black line is fit to population A, while
  the dashed black line is fit to population E.}
\end{figure*}

\begin{table*}
  \centering
  \begin{tabular}{c | c c c c c c}
    \hline
    Population & Age & Distance Modulus & Extinction & Y & $\alpha_{ML}$ & $\chi^{2}_{\nu}$\\
    & [Gyr] & & [mag] & & &\\
    \hline
    \hline
    A & 12.996$^{+0.87}_{-0.64}$ & 15.021 & 0.54 & 0.24 & 2.050 & 0.021\\
    E & 13.061$^{+0.86}_{-0.69}$ & 15.007 & 0.537 & 0.39 & 1.600 & 0.033 \\
    \hline
  \end{tabular}
  \label{tab:BestFitResults}
  \caption{Best fit parameters derived from fitting isochrones to the fiducual lines derived from the NCG 2808 photometry. The one sigma uncertainty reported on population age were determined from the 16th and 84th percentiles of the distribution of best fit isochrones ages.}
\end{table*}

@@

\subsection{The Number of Populartions in NGC 2808}
In order to estimate the number of populations which ideally fit the NGC 2808
F275W-F814W photometry without overfitting the data we make use of silhouette
analysis \citep[][and in a similar manner to how \citet{Valle2022} preform
their analysis of spectroscopic data]{ROUSSEEUW198753}. We find the average
silhouette score for all tagged clusters identified using BGMM in all magnitude
bins over the CMD using the standar python module \texttt{sklearn}. Figure
\ref{fig:clusterAn} shows the silhouette analysis results and that two
populations fit the photometry most ideally. This is in line with what our BGMM
model predicts for the majority of the the CMD.
